---
title: "Tuberculosis In Wuhan China"
author: "Ogechi Daniel Koel"
format: docx
editor: visual
---

## Introduction

Tuberculosis (TB) is an infectious disease caused by the bacterium *Mycobacterium tuberculosis*. It primarily affects the lungs but can also spread to other parts of the body, such as the kidneys, spine, and brain. TB spreads through the air when an infected person coughs, sneezes, or talks, making it highly contagious2.

This dataset contains monthly tuberculosis (TB) incidence per 100,000 individuals in a town in China from 2012 to 2019. I will develop an ARIMA/SARIMA model to forecast TB incidence for the next three years, ensuring that seasonal patterns and long-term trends are effectively captured.

```{r warning=FALSE, message=FALSE}
#loading the libraries 
library(here)
library(tidyverse)
library(readxl)
library(tseries)
library(forecast)
library(flextable)
```

```{r warning=FALSE, message=FALSE}
#loading the dataset
data <- read_excel(here("data", "data.xlsx"))
```

DATA EXPLORATION

```{r warning=FALSE, message=FALSE}
#viewing the first 10 rows

head(data, 10)


```

```{r warning=FALSE, message=FALSE}
#viewing the last 10 rows 

tail(data, 10)

```

Our data is univariate, has one variable that varies with time thus suitable for SARIMA or ARIMA model.

```{r warning=FALSE, message=FALSE}
#checking for missing values 

sum(is.na(data))
```

```{r warning=FALSE, message=FALSE}
#starting time 
min(data$Time)

#ending time 
max(data$Time)

```

The TB incidence starts from January of 2012 and ends in June of 2019.

```{r warning=FALSE, message=FALSE}
#checking for the structure of the data 
str(data)
```

The dataset is not in time series format and therefore needs to be converted into a `ts` format.

```{r warning=FALSE, message=FALSE}
#convert the data to time series format

data <- ts(data$`TB incidence（per 100,000 populations）`, 
           start =c(2012, 01), end = c(2019, 06), frequency = 12)
#checking is the data is now a time series.
attributes(data)

```

The TB incidence is recorded monthly, which is why I used a frequency of 12.

[Plotting the Data]{.underline}

```{r warning=FALSE, message=FALSE}
#plotting the time series 
plot(data, main = "Time series")

```

From the plot, TB incidence decreases with time with some seasonal variations.

```{r warning=FALSE, message=FALSE}
plot(aggregate(data, FUN = mean), ylab = "TB Incidence",
     main = "TB incidence General Trend")
```

TB incidence decreases over time possibly indicating better adherence to regimens by the patients or implementation of effective policies by the government.

```{r warning=FALSE, message=FALSE}

#decomposing to see if there is seasonality

decomposed <- decompose(data,  type = "additive")



#plotting
plot(decomposed)
```

The seasonal plot distinctly displays repeating patterns at regular intervals, which is a clear sign of seasonality.

```{r warning=FALSE, message=FALSE}
# Seasonal plot
ggseasonplot(data, year.labels = TRUE, main = "Seasonal Patterns in TB Incidence")+
  theme_bw()
```

There is a high TB incidence in march over the Years possibly because march marks the transition from winter to spring in China, when respiratory infections tend to spike due to colder temperatures and increased indoor crowding. This could contribute to higher TB transmission.

[**FITTING THE SARIMA MODE**]{.underline}

```{r warning=FALSE, message=FALSE}
#Fitting the model

fit <- auto.arima(data, stationary = FALSE, seasonal =TRUE)

```

I will be using SARIMA model since TB incidence exhibited some seasonality. I'm using auto.arima() that fits the automatically fits the model using AICc.

```{r warning=FALSE, message=FALSE}
#checking if it was SARIMA that was fitted
fit
```

[**MODEL DIAGNOSTICS**]{.underline}

a\) Normality Of residuals

```{r warning=FALSE, message=FALSE}
hist(residuals(fit), col = "skyblue", main = "Histogram of the residuals",
     xlab = "Residuals")
```

The residuals appears to be roughly normally distributed.

b). Homoscedasticity

```{r warning=FALSE, message=FALSE}
#Homoscedasticity (mean = 0 and variance = 1)
plot(residuals(fit), main = "Residuals over Time")
abline(h = 0, col = "red")
```

The residuals appear to be roughly constant around the mean 0 thus homoscedasticity of the residuals holds.

c\) Autocorrelation

i\. Ljung Box test

```{r warning=FALSE, message=FALSE}
Box.test(residuals(fit))
```

We fail to reject the null hypothesis (p=0.9997). The residuals are therefore independently distributed.

ii\) ACF plot

```{r warning=FALSE, message=FALSE}
acf(residuals(fit), main = "ACF plot of the residuals")
```

The residuals are not correlated because the autocorrelations fall within the confidence bars.

-Our model is fit for forecasting.

```{r warning=FALSE, message=FALSE}
#Forecating for three years
forecasted <- forecast(fit, level = c(95), h = 36) 
forecasted
```

The forecasted values tend to decrease over time, and the narrow confidence intervals suggest greater reliability and significance..

```{r warning=FALSE, message=FALSE}
#plotting the forecasted values
plot(forecasted, main = "Forecasted values")
```

The forecated values follow the suit (decreasing trend) as shown in the above plot.
